### 2014-11-05 Introduction.tex
lots of modifications

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing these systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking, where abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.  There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

@@

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.
We are able to verify
  the several components of a modern processor 
\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

>>>>

The goal of this work is to provide a framework for full verification of such
complex hardware systems. Hardware designers already recognize the challenge of
implementing these systems, to such an extent that they have already become some
of the most serious real-world adopters of formal methods.  Hardware
verification is dominated by model-checking, where abstraction techniques are
used to reduce designs to finite state spaces, which can be explored
exhaustively.  There are limits to the construction of sound abstractions, so
verifications of, e.g., cache-coherence protocols have almost always treated
systems with concrete topologies, involving particular finite numbers of caches
and processors. For instance, explicit-state model checking tools like
Murphi~\cite{murphi} or TLC~\cite{tlc} are only able to handle small and
simplified systems, \eg{} a single-level cache hierarchy with three CPUs, two
addresses, and one bit of data. Symbolic model checking techniques have fared
slightly better: McMillan \etal{} have verified a 2-level MSI protocol based on
the Gigamax distributed multiprocessor using SMV~\cite{gigamax}. Optimizations
on these techniques (\eg{} partial order reduction~\cite{part}, symmetry
reduction~\cite{sym1, sym2}, compositional reasoning~\cite{comp, Mccomp, mcc}, \etc{}) scale
the approach further, but they are still unable to handle realistic
hierarchical protocols. We might hope to achieve higher assurance and
understanding of design ideas by verifying \emph{infinite families} of hardware
designs, which resist reduction to finite-state models.  A potential related
benefit is reducing the computational resources required to formally verify a
new design in the family, applying a more general theorem without much new
analysis.

@@

\item A general \emph{modular verification methodology} applicable to hardware
  algorithms, based on \emph{labeled transition systems}.
We are able to verify
  the several components of a modern processor 
%\adam{(Figure~\ref{zoom} OK to have a forward reference to this figure?}
against general formal interfaces,
  enabling mixing and matching of different realizations of each interface,
  without doing any new proofs that peek beneath abstraction boundaries.

### 2014-11-05 TransitionsUsefulSecond.tex
lots of modifications

We can define sequential consistency using the simple memory from Figure
\ref{$M_m$} as follows, for $n$ processors:
\begin{defn}
$SC \triangleq P^n_\text{ref} + M_m$
\label{sc}
\end{defn}

@@

Figure~\ref{Ooo} defines that LTS,
P$_{\texttt{so}}$. This processor may issue arbitrary speculative
loads, but it only ever \emph{commits} the next instruction in serial
program order. 
We will see the processor issue two kinds of loads, a speculative load (with
the label suffix $\Spec$) and a commit or real load (with the label suffix $\Real$).  To
maintain SC, every speculative
load must have a matching verification load later on, and we maintain
the illusion that the program only depends on the results of
verification loads, which, along with stores, \emph{must be issued in
serial program order}.

>>>>

We can define sequential consistency using the simple memory from Figure
\ref{M_m} as follows, for $n$ processors:
\begin{defn}
$SC \triangleq P^n_\text{ref} + M_m$
\label{sc}
\end{defn}

@@

Figure~\ref{Ooo} defines that LTS, P$_{\texttt{so}}$. This processor may issue
arbitrary speculative loads (with label suffix $\Spec$), but it only ever
\emph{commits} the next instruction in serial program order. These loads are
issued non-deterministically from any load instruction after its address is
computed.  This non-deterministic issue is going to be necessary when the
memory LTS $M_m$ is replaced by caches.  To maintain SC, every speculative load
must have a matching verification load later on (with label suffix $\Real$),
and we maintain the illusion that the program only depends on the results of
verification loads, which, along with stores, \emph{must be issued in serial
program order}.

### 2014-11-05 TransitionsUsefulThird1.tex
lots of modifications

Fields $\dt, \s$ are maps over pairs of cache names and memory addresses.
The $\dt$ map assigns each pair a memory-cell value, and the $\s$ map
assigns each pair a coherence state.  A coherence state is $M, S$ or
$I$, broadly representing permissions to modify, read, or do nothing
with an address, respectively, the decreasing permissions denoted by $M > S > I$. This interpretation of permissions is
accurate at the leaf caches connected to the processors (\ie{} L1 caches),
but the interpretation is subtly different at the higher level caches.

Mapping $\wt$ goes from cache name and address to either $\epsilon$ or a
coherence state.  It denotes what permission the cache is waiting for,
if any.  That is, a cache may have decided to \emph{upgrade} its
coherence state for some address to a more permissive value, but the
cache is waiting for acknowledgment from another cache before
finalizing the upgrade.

Since we have a directory-based protocol, the $\s$ and
$\wt$ maps have mirror-image counterparts $\dst$ and $\dwt$.
The former maps triples of parent name, child name, and address to
coherence state. It represents the parent's notion of the
coherence state of the child for that address. We later prove that this notion
is always conservative, \ie{} if the parent assumes that a child does not have
a particular permission, then it is guaranteed in this system that the child
will not have that permission.  The $\dwt$ field also represents a mapping from parent
name, child name, and address to either $\epsilon$ or coherence state. It
denotes whether the parent is waiting for any downgrade response from the
corresponding child, and if so, the coherence state that the child must
downgrade to as well.

All communication channels in the system are represented by field $\ch$.
The channels are asymmetric as
follows. Channels going from parent to children (indexed by the names of the
parent and child) maintain FIFO order of messages between the same parent-child
pair. They carry both request messages (asking the child to downgrade) and
response messages (acknowledging that the child can upgrade). Channels going
from children to their parents are further split into request and response
channels, thus indexed by tuples (child name, parent name, request or response).
These channels need not maintain the FIFO order for messages between the same
child-parent pair (notice the use of $\coloncolon$ to add and remove elements
from the parent to child channel, as opposed to $\uplus$ for the 
child to parent channel). The reason is that only one request or one response is
sent for a particular address between a child-parent pair -- this invariant can
be proved for the LTS in  Figure \ref{cache}. The order between a request and
response message sent from the same child to its parent is lost, but our
protocol ensures that nothing bad happens as a result.  Notice rules
ParentRecvRsp and ParentRecvReq in Figure \ref{cache}, each of which looks at the
previous coherence state of the incoming message and compares it with its own
directory state to decide whether to process the message.

Here is an intuition on how the transitions work in the common case.  A cache
can decide, spontaneously, to upgrade its coherence state, in which case it
sends an upgrade request to its parent. The parent then makes a local decision
on whether to send a response to the requesting child or not, based on its
directory approximation and its own coherence state $\s$. If $\s$ is lower than
the requested upgrade, then it cannot handle the request, and instead must
decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
upgrade, it makes sure that the rest of its children are ``compatible'' with
the requested upgrade (given by $\compat$ definition below).  If not, the
parent must send requests to the incompatible children to downgrade. Finally, when
the $\s$'s upgrade and children's downgrade responses are all received, the
original request can be responded to. The transitions for sending requests are
allowed to happen spontaneously without any trigger. An L1 cache can process a
load request or a store request only if it is in at least the \Sh{} state or
\Mo{} state, respectively, for the requested address; otherwise it must
spontaneously request to upgrade.

@@

\section{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
\label{safety}

@@

The proof that $M_c$ obeys store atomicity is involved enough to deserve a
section's worth of explanation, which we now turn to.

We first reduce the proof of store atomicity to several lemmas.
We believe that other invalidation-based
protocols with inclusive caches (\ie{} those cache hierarchies where a parent
cache contains all the locations that any of its children contains) like MESI,
MOSI, and MOESI also satisfy our lemmas, making their verification
easier by reusing most of our MSI proof; we omit the discussion of extension of
this proof to other protocols for space reasons.

The lemmas
presented below hold after any number of transitions
starting from the initial state.

While we have proofs in Coq for each lemma, our discussion will be mainly
informal for brevity. All the proofs can be found in our supplementary material.

Intuitively our protocol is correct because it is guaranteed that any cache that attempts to
handle a load request has the ``latest'' value.

@@

It is easy to see that any two states $y_1 \le x_1$ and $y_2 \le x_2$ are also
compatible if $x_1$ and $x_2$ are compatible.

@@

Lemmas~\ref{conservative} and~\ref{localCompatibility1} imply a
$\ge$ relation between the states of a parent and its child for any address.
Since the ancestor relation is the reflexive-transitive closure of the parent
relation, the $\ge$ relation is preserved between the states of two caches $p$
and $c$ where $p$ is the ancestor of $c$.

Any two caches $c_1, c_2$ neither of which is an ancestor of the other have a
lowest common ancestor $x$, which has
two distinct children $x_1$ and $x_2$ such that they are the ancestors of $c_1$
and $c_2$ respectively. For any address $a$, $\s(x_1, a)$ and
$\s(x_2, a)$ are compatible by Lemma~\ref{localCompatibility2}. From
the previous paragraph about the relation between the states of a cache and its
ancestor, it follows that $c_1$ and $c_2$ are also compatible, proving
Lemma~\ref{nonAncestorCompatible}.

Using Lemmas~\ref{nonAncestorCompatible} and~\ref{noTransitWrite}, one can
prove Lemma~\ref{latestValue}. The proof is by induction on time $T$; we assume
that any \clean{} (node, address) pair has the latest value until time
$T$. Consider a (node, addr) pair at time $T+1$.
If the pair was \clean{} at $T$, then no cache,
save itself (because of Lemma~\ref{nonAncestorCompatible}), can process any
store request for that address, thus continuing to be \clean{} and having the latest value.
And if the pair was not \clean{} at time $T$, it will receive data from a cache
that is \clean{} for the (node, address) pair, and no store
request can be processed when the data is in transit (by
Lemma~\ref{noTransitWrite}). Therefore, by the induction hypothesis the cache will
continue to have the latest value whenever it is \clean{} for an address at time $T+1$.

Analyzing a parent-child pair will suffice in proving
Lemmas~\ref{conservative}, \ref{localCompatibility1} and \ref{localCompatibility2}.  Proof of
Lemma~\ref{noTransitWrite} is similar to the proof of
Lemma~\ref{nonAncestorCompatible}.

>>>>

A coherence state is $M, S$ or $I$, broadly representing permissions to modify,
read, or do nothing with an address, respectively, the decreasing permissions
denoted by $M > S > I$. This interpretation of permissions is accurate at the
leaf caches connected to the processors, but the
interpretation is subtly different at the higher level caches.

$\wt$ denotes what permission the cache is waiting for,
if any.  That is, a cache has decided to \emph{upgrade} its
coherence state for some address to a more permissive value, but
it is waiting for acknowledgment from its parent before
finalizing the upgrade.

$\dst$ represents the parent's notion of the
coherence state of the child for that address. We later prove that this notion
is always conservative, \ie{} if the parent assumes that a child does not have
a particular permission, then it is guaranteed in this system that the child
will not have that permission.  $\dwt$
denotes whether the parent is waiting for any downgrade response from the
corresponding child, and if so, the coherence state that the child must
downgrade to as well.

All communication channels in the system are represented by field $\ch$.  The
channels are asymmetric as follows. Channels going from parent to children
(indexed by the names of the parent and child) maintain FIFO order of messages
between the same parent-child pair. They carry both request messages (asking
the child to downgrade) and response messages (acknowledging that the child can
upgrade). Channels going from children to their parents are further split into
request and response channels, thus indexed by tuples (child name, parent name,
request or response).  These channels need not maintain the FIFO order for
messages between the same child-parent pair (notice the use of $\coloncolon$ to
add and remove elements from the parent to child channel, as opposed to
$\uplus$ for the child to parent channel). The reason is that only one request
or one response is sent for a particular address between a child-parent pair --
this invariant can be proved for the LTS in  Figure \ref{cache}. The order
between a request and response message sent from the same child to its parent
is lost, but our protocol ensures that they are processed in order by the parent, by
comparing the previous coherence state of the incoming messages with the
parent's directory state (in rules ParentRecvRsp
and ParentRecvReq).

Here is an intuition on how the transitions work in the common case.  A cache
can decide, spontaneously, to upgrade its coherence state, in which case it
sends an upgrade request to its parent. The parent then makes a local decision
on whether to send a response to the requesting child or not, based on its
directory approximation and its own coherence state $\s$. If $\s$ is lower than
the requested upgrade, then it cannot handle the request, and instead must
decide to upgrade $\s$.  Once the parent's $\s$ is not lower than the requested
upgrade, it makes sure that the rest of its children are ``compatible'' with
the requested upgrade (given by $\compat$ definition below).  If not, the
parent must send requests to the incompatible children to downgrade. Finally, when
the $\s$'s upgrade and children's downgrade responses are all received, the
original request can be responded to. The transitions for sending requests are
allowed to happen spontaneously without any trigger. $P_\text{ref}$ issues speculative
loads only for addresses which are in \Sh{} or \Mo{} state in L1; otherwise L1
spontaneously requests an upgrade for that address.

@@

\subsection{Proof that the MSI Protocol Preserves Store Atomicity}\label{sec:ccproof}
\label{safety}

@@

The proof that $M_c$ obeys store atomicity is involved enough that we will only
state the lemmas used to prove this. The proofs in Coq for each theorem and
lemma in this section can be found in our supplementary material.

We believe that other invalidation-based protocols with inclusive caches (\ie{}
those cache hierarchies where a parent cache contains all the locations that
any of its children contains) like MESI, MOSI, and MOESI also satisfy our
lemmas, making their verification easier by reusing most of our MSI proof; we
omit the discussion of extension of this proof to other protocols for space
reasons.

The following lemmas hold in the state reached after any number of transitions
starting from the initial state.

@@

@@
